Neural Circuit Organization of the Ctenophore Apical Organ and Its Role in Gravity Sensing

### 2025-03-03 manuscript/Manuscript.qmd

---
title: "Synaptic connectome of the ctenophore apical organ"

@@

**text in bold** *italic* [underline]{.underline}

@@

![**Figure 3. Reconstruction of Balancer Ciliated Cells in the Apical Organ**\
(A) Connectivity matrix of AO neurons. Columns represent presynaptic AO neurons, rows represent postsynaptic AO neurons, and each number indicates the number of synapses. AO neurons are interconnected via synapses, and AO neuron Q1234 forms a self-synapse, autapse.\
(B) Three-dimensional reconstruction of balancer ciliated cells in all four quadrant regions. Each quadrant contains approximately 30 cells. All balancer cilia are monocilia. The spheres indicate the positions of the nuclei. The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(C) Synaptic connections from AO neurons to balancer ciliated cells. The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view. Balancer ciliated cells are depicted in light gray, AO neuron Q1234 in black, AO neuron Q12 in orange, and AO neuron Q34 in blue.\
(D) Connectivity map of three AO neurons (Q1234 in blue, Q12 in pink, and Q34 in orange) and balancer ciliated cells (light blue). The thickness of the arrows and the numbers correspond to the number of synapses. Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.](figures/Figure3.png)

@@

![**Figure 4. Reconstruction of bridge cells and gravity-sensing neural circuits in the apical organ**\
(A) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4 quadrants. The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color. The morphology of individual bridge cells extending across opposite quadrant regions is depicted. The spheres represent the positions of individual nuclei. The left panel shows an aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(B) Mitochondrial localization within bridge cells and associated presynaptic triad structures. Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified. The left panel shows a dorsal view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(C) Synaptic connections between AO neurons and bridge cells. The positions of synapses from AO neurons to bridge cells (red) and from bridge cells to AO neurons (light blue) are indicated. The left panel shows an aboral view, while the right panel presents a tentacular plane lateral view. Bridge cells are shown in light gray, AO neurons in Q1234 are in black, AO neurons in Q12 are in orange, and AO neurons in Q34 are in blue.\
(D) Connectivity matrix of the gravity-sensing neural circuit. Columns represent presynaptic cell groups, while rows represent postsynaptic cell groups. The numbers and varying shades of blue correspond to the number of synapses.\
(E) Complete connectivity map of the gravity-sensing neural circuit. Cells belonging to the same group are enclosed in hexagons, and the number of cells is added to their labels. The thickness of the arrows and the numerical values indicate the number of synapses. AO neurons in Q1234 are shown in black, AO neurons in Q12 in yellow, AO neurons in Q34 in blue, balancer cell groups in gray, and bridge cell groups in pink.](figures/Figure4.png)

@@

![**Figure 5. Differences in Balancer Cilia Movement Control Due to Variations in the Gravity-Sensing Neural Circuit Pathways**\
(A) Schematic diagram of differential interference contrast (DIC) microscopy setup for imaging balancer cilia movement. The microscope was tilted 90 degrees so that the stage was positioned vertically. A monochrome CMOS camera sensitive to near-infrared (NIR) light was used, synchronized with an 850 nm strobe light source. The movement of the balancer cilia was recorded using a 40× objective lens.\
(B, C) Sagittal plane view (B) and tentacular plane view (C) of the apical organ under the microscope. The direction of gravity is always oriented downward in the images. The two balancer cilia being compared are indicated by arrowheads.\
(D, E) Connectivity maps of the gravity-sensing neural circuit corresponding to Figure 3E. These maps emphasize the differences in neural circuits in the sagittal plane (D) and tentacular plane (E), focusing on the balancer cell groups (blue hexagons). AO neuron Q1234 is shown in green, while AO neurons Q12 and Q34 are shown in pink.\
(F) Kymograph patterns comparing the arrest and re-beat of balancer cilia movement. The left and right balancer cilia movements are compared in both the sagittal plane (left) and tentacular plane (right). The direction and length of the arrows indicate time.\
(G) Graph showing the time differences in the arrest and re-beat timing of the left and right balancer cilia. S and T represent the sagittal plane and tentacular plane, respectively.](figures/Figure5.png)

Figure 6 - cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.

### Synapses

"synaptic regions with the characteristic presynaptic triad mor- phology of ctenophore nerves (i.e., a single layer of vesicles, smooth ER sac, and closely apposed mito- chondria [@Horridge_1964; @Hernandez_Nicaise_1973] "

Figure 7 - synapses, reconstruction, annotation, numbers, distribution of partners, example EM (figure supplement - lots of EM), mitochondria (Fig suppl - mitochondrion stats - Sanja)

Gap juntions? Innexins are expressed in the apical organ in specific patches of cells in each quadrant [@Ortiz_2023]

Ctenophore synapses [@Hernandez_Nicaise_1973]

We found no synapses between bridge and balancer, in agreement with Tamm [@Tamm_2002].

Figure 8 - connectome cell-type based, quadrant based, nerve net syntycia, bridge, quadrants (Fig suppl - cell-based network)

Tamm&Tamm observed synapses from neurites to bridge cells (2002). and in Mnemiopsis larvae observed synapses from bridge cells onto other cells

Inserting Figures

You can add your figures into the rendered document. We saved the figures into /manuscript/figures or /manuscript/figure_supplements and can insert them from there. We use knitr::include_graphics for this. The title and legend can also be edited, as will as the width of the output figure.

### Connectivity

### Physiology?

@@

![**Figure 6. Comparison of Neural Circuits Regulating Ciliary Movement in *M. leidyi* (Ctenophore) and *P. dumerilii* (Annelid)**\
Each neuron represented by a circle is color-coded to indicate putative homologous cell types involved in ciliary movement regulation. Ciliated cells are shown in gray. Synapses are indicated by arrows, with magenta representing synapses that induce ciliary arrest and blue representing synapses that induce ciliary re-beat or an increase in ciliary beating frequency.\
(A) Gravity-sensing neural circuit of *M. leidyi*, as revealed in this study. Bridge cells (yellow squares) are suggested to be electrically coupled (indicated by yellow zigzag lines), implying their potential involvement in feedback mechanisms between neurons and ciliated cells.\
(B) Neural circuit regulating prototroch ciliary movement in *P. dumerilii*. Serotonergic neurons (Ser-h1 neurons, blue) activate ciliary movement, while cholinergic neurons (MC neurons, magenta) induce ciliary arrest. Arrows directed toward ciliated cells in the head prototroch are color-coded accordingly.](figures/Figure6.png)

>>>>

---
title: "Neural Circuit Organization of the Ctenophore Apical Organ and Its Role in Gravity Sensing"

@@

Understanding the evolution and function of neurons is crucial for elucidating the origins of animal behavior. Ctenophores, as the earliest-diverging lineage with a nervous system, provide a unique model for investigating neural circuit organization. In this study, we performed volumetric electron microscopy (EM) reconstruction of the apical organ in Mnemiopsis leidyi to characterize its neural network involved in gravity sensing. We reconstructed 909 cells, including novel syncytial neurons spanning multiple quadrants, which form complex synaptic networks with balancer cells—key mechanosensory components controlling ciliary movement. Our connectome analysis revealed that these syncytial neurons establish direct and indirect pathways to balancer cells, with "bridge cells" acting as afferent relay neurons. High-speed imaging of balancer cilia movements further demonstrated differential regulation between sagittal and tentacular planes, suggesting distinct functional roles of apical organ neurons in modulating posture and locomotion. These findings provide new insights into the structure and function of ctenophore neural circuits, shedding light on early nervous system evolution and mechanosensory integration in basal metazoans.

@@

@@

![**Figure 3. Reconstruction of Balancer Ciliated Cells in the Apical Organ Reconstruction of bridge cells and gravity-sensing neural circuits in the apical organ**\
(A) Connectivity matrix of AO neurons. Columns represent presynaptic AO neurons, rows represent postsynaptic AO neurons, and each number indicates the number of synapses. AO neurons are interconnected via synapses, and AO neuron Q1234 forms a self-synapse, autapse.\
(B) Three-dimensional reconstruction of balancer ciliated cells in all four quadrant regions. Each quadrant contains approximately 30 cells. All balancer cilia are monocilia. The spheres indicate the positions of the nuclei. The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(C) Synaptic connections from AO neurons to balancer ciliated cells. The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view. Balancer ciliated cells are depicted in light gray, AO neuron Q1234 in black, AO neuron Q12 in orange, and AO neuron Q34 in blue.\
(D) Connectivity map of three AO neurons (Q1234 in blue, Q12 in pink, and Q34 in orange) and balancer ciliated cells (light blue). The thickness of the arrows and the numbers correspond to the number of synapses. Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.
(A) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4 quadrants. The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color. The morphology of individual bridge cells extending across opposite quadrant regions is depicted. The spheres represent the positions of individual nuclei. The left panel shows an aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(B) Mitochondrial localization within bridge cells and associated presynaptic triad structures. Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified. The left panel shows a dorsal view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(C) Synaptic connections between AO neurons and bridge cells. The positions of synapses from AO neurons to bridge cells (red) and from bridge cells to AO neurons (light blue) are indicated. The left panel shows an aboral view, while the right panel presents a tentacular plane lateral view. Bridge cells are shown in light gray, AO neurons in Q1234 are in black, AO neurons in Q12 are in orange, and AO neurons in Q34 are in blue.\
(D) Connectivity matrix of the gravity-sensing neural circuit. Columns represent presynaptic cell groups, while rows represent postsynaptic cell groups. The numbers and varying shades of blue correspond to the number of synapses.\
(E) Complete connectivity map of the gravity-sensing neural circuit. Cells belonging to the same group are enclosed in hexagons, and the number of cells is added to their labels. The thickness of the arrows and the numerical values indicate the number of synapses. AO neurons in Q1234 are shown in black, AO neurons in Q12 in yellow, AO neurons in Q34 in blue, balancer cell groups in gray, and bridge cell groups in pink.](figures/Figure3.png)

@@

![**Figure 4. Differences in Balancer Cilia Movement Control Due to Variations in the Gravity-Sensing Neural Circuit Pathways Comparison of Neural Circuits Regulating Ciliary Movement in *M. leidyi* (Ctenophore) and *P. dumerilii* (Annelid)**\
(A) Schematic diagram of differential interference contrast (DIC) microscopy setup for imaging balancer cilia movement. The microscope was tilted 90 degrees so that the stage was positioned vertically. A monochrome CMOS camera sensitive to near-infrared (NIR) light was used, synchronized with an 850 nm strobe light source. The movement of the balancer cilia was recorded using a 40× objective lens.\
(B, C) Sagittal plane view (B) and tentacular plane view (C) of the apical organ under the microscope. The direction of gravity is always oriented downward in the images. The two balancer cilia being compared are indicated by arrowheads.\
(D, E) Connectivity maps of the gravity-sensing neural circuit corresponding to Figure 3E. These maps emphasize the differences in neural circuits in the sagittal plane (D) and tentacular plane (E), focusing on the balancer cell groups (blue hexagons). AO neuron Q1234 is shown in green, while AO neurons Q12 and Q34 are shown in pink.\
(F) Kymograph patterns comparing the arrest and re-beat of balancer cilia movement. The left and right balancer cilia movements are compared in both the sagittal plane (left) and tentacular plane (right). The direction and length of the arrows indicate time.\
(G) Graph showing the time differences in the arrest and re-beat timing of the left and right balancer cilia. S and T represent the sagittal plane and tentacular plane, respectively.
Each neuron represented by a circle is color-coded to indicate putative homologous cell types involved in ciliary movement regulation. Ciliated cells are shown in gray. Synapses are indicated by arrows, with magenta representing synapses that induce ciliary arrest and blue representing synapses that induce ciliary re-beat or an increase in ciliary beating frequency.\
(A) Gravity-sensing neural circuit of *M. leidyi*, as revealed in this study. Bridge cells (yellow squares) are suggested to be electrically coupled (indicated by yellow zigzag lines), implying their potential involvement in feedback mechanisms between neurons and ciliated cells.\
(B) Neural circuit regulating prototroch ciliary movement in *P. dumerilii*. Serotonergic neurons (Ser-h1 neurons, blue) activate ciliary movement, while cholinergic neurons (MC neurons, magenta) induce ciliary arrest. Arrows directed toward ciliated cells in the head prototroch are color-coded accordingly.](figures/Figure4.png)

@@
